Verified Nanopasses for Compiling Conditionals

### 2025-08-21 paper/paper.tex
fixes

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigplan,review,dvipsnames,screen,10pt]{acmart}
%\documentclass[sigplan,screen]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[OlivierFest '25]{Workshop dedicated to Olivier Danvy's 65th birthday}{October,
  2025}{Singapore}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Verified Nanopasses for Compiling Conditionals}

@@

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
%% \begin{CCSXML}
%% <ccs2012>
%%  <concept>
%%   <concept_id>00000000.0000000.0000000</concept_id>
%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%%   <concept_significance>500</concept_significance>
%%  </concept>
%%  <concept>
%%   <concept_id>00000000.00000000.00000000</concept_id>
%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%%   <concept_significance>300</concept_significance>
%%  </concept>
%%  <concept>
%%   <concept_id>00000000.00000000.00000000</concept_id>
%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%%   <concept_significance>100</concept_significance>
%%  </concept>
%%  <concept>
%%   <concept_id>00000000.00000000.00000000</concept_id>
%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%%   <concept_significance>100</concept_significance>
%%  </concept>
%% </ccs2012>
%% \end{CCSXML}

%% \ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
%% \ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
%% \ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
%% \ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{short-cut boolean evaluation, nanopass, verified compilation, mechanized proof, Agda}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
%% \begin{teaserfigure}
%%   \includegraphics[width=\textwidth]{sampleteaser}
%%   \caption{Seattle Mariners at Spring Training, 2010.}
%%   \Description{Enjoying the baseball game from the third-base
%%   seats. Ichiro Suzuki preparing to bat.}
%%   \label{fig:teaser}
%% \end{teaserfigure}

@@

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to an algorithm of
\citet{Aho:1986qf} (presented in the form of an attribute grammar)
that translates Boolean expressions and conditionals into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarrassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
algebraic datatypes~\citep{PEYTONJONES19983}.

@@

Towards developing a course on verified compilation, I am interested
in proving correctness of the compiler presented in the
\emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. While it does not include
proofs of correctness for its compiler passes, the textbook is
organized in a way that might facilitate the proofs of correctness.
%
\emph{Essentials of Compilation} takes an \emph{incremental} approach
to teaching compilers, that is, it guides the students through the
creation of multiple compilers for a sequence of languages, each of
which adds a language feature to the previous language. For the
language with Boolean expressions and conditionals, the Explicate
Control pass of the compiler uses the one-function per context
approach of Danvy to translate into an x86-flavored register transfer
language.
%
In addition to being incremental, \emph{Essentials of Compilation}
employs the \emph{nanopass} approach, breaking up the compiler into
many small passes that are easier for students to
digest~\citep{Sarkar:2004fk}.
%
There are many other pedagogical treatments of
compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa}
but none of them take the incremental approach.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The Agda code is available in the supplementary
material and at the following location.
\begin{center}
\url{https://github.com/jsiek/correct_compilers}
\end{center}
The Agda code does not depend on special features of Agda and could
have been carried out in Rocq, Isabelle, or Lean.\footnote{The choice
of Agda for this paper is not an endorsement of Agda but is a
byproduct of the author's relative expertise in Agda at the time of
writing.}

@@

\subsection{The C-like language}
\label{sec:c-lang}

@@

\subsection{x86 with variables}

@@

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    let x = read in
    let y = read in
    if (if x == 0
        then true
        else read == 1)
    then y - 1
    else (y - 2) - x
\end{lstlisting}
  \hrule
\caption{An example source program.}
\label{fig:running-example-source}
\end{figure}

@@

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    let x = read in
    let y = read in
    if (if x == 0
        then true
        else (let t1 = read in t1 == 1))
    then (y - 1)
    else (let t2 = y - 2 in t2 - x)
\end{lstlisting}
  \hrule
\caption{\small Result of Remove Complex Operands on the example.}
\label{fig:running-example-monadic}
\end{figure}

@@

\subsubsection{Remove Complex Operands is Correct}

@@

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    var x, y, t1, t2;
    x := read;
    y := read;
    if (if x == 0
        then true
        else (t1 := read; t1 == 1)
    then (y - 1)
    else (t2 := y - 2; t2 - x)
\end{lstlisting}
  \hrule
  \caption{Result of Lift Locals on the example.}
  \label{fig:running-example-structured-imperative}
\end{figure}
The \lstinline{lift-locals} function delegates to the recursive
function \lstinline{lift-locals-mon}, where all the work is done.
That function returns (1) a number $n$ that counts the number of
\lstinline{let} bindings that were in the expression and (2) the
translation of the expression.

@@

\subsubsection{Lift Locals is Correct}

@@

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    var x, y, t1, t2;
    block_0:
       x := read;
       y := read;
       if (x == 0) goto block_1; else goto block_3;
    block_1:
        goto block_2;
    block_2:
        return y - 1;
    block_3:
       t1 := read;
       if (t1 == 1) goto block_4; else goto block_5;
    block_4:
        goto block_2;
    block_5:
        goto block_6;
    block_6:
        t2 := y - 2;
        return t2 - x;
\end{lstlisting}
  \hrule
  \caption{Example after Explicate Control, where control flow
    is made explicit with goto.}
\label{fig:running-example-goto}
\end{figure}

@@

For the finale of the Explicate Control pass, we discuss the
translation of the conditional \lstinline{(If e$₁$ e$₂$ e$₃$)}
in \lstinline{explicate-pred}.  The two
branches \lstinline{e$₂$} and \lstinline{e$₃$} are in predicate
position, so we compile them with recursive calls to
\lstinline{explicate-pred} to obtain \lstinline{c$₂$} and
\lstinline{c$₃$}. However, they share the same two continuations,
\lstinline{thn} and \lstinline{els}, so we add both of them to the CFG
to avoid code duplication. We compile \lstinline{e$₁$} with
\lstinline{explicate-pred}, passing \lstinline{c$₂$} and
\lstinline{c$₃$} as the continuations.

\begin{lstlisting}
explicate-pred (If e$₁$ e$₂$ e$₃$) thn els =
   add-node thn thenB
   $λ$ l$₁$ $→$ add-node els thenB
   $λ$ l$₂$ $→$ explicate-pred e$₂$ (Goto l$₁$) (Goto l$₂$) thenB
   $λ$ c$₂$ $→$ (explicate-pred e$₃$ (Goto l$₁$) (Goto l$₂$)) thenB
   $λ$ c$₃$ $→$ explicate-pred e$₁$ c$₂$ c$₃$
\end{lstlisting}

@@

\begin{lstlisting}
explicate-pred-correct :
  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G G$′$}
  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
  $→$ ((s$′$,$ρ′$) , G $⊢ᶜ$ (if b then c$₂$ else c$₃$) $⇓$ v $⊣$ (s$″$,$ρ″$))
  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G $≡$ (c$₁$ , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
\end{lstlisting}

@@

\begin{figure}[tp]
  \hrule
\begin{lstlisting}[mathescape=false,basicstyle=\ttfamily\footnotesize]
    var x, y, t1, t2
    block_0:
       callq read_int
       movq %rax, x
       callq read_int
       movq %rax, y
       cmpq x, $0
       je block_1
       jmp block_3
    block_1:
        jmp block_2
    block_2:
        movq y, %rax
        subq $1, %rax
        movq %rax, %rax
    block_3:
        callq read_int
        movq %rax, t1
        cmpq t1, $1
        je block_4
        jmp block_5
    block_4:
        jmp block_2
    block_5:
        jmp block_6
    block_6:
        movq y, %rax
        subq $2, %rax
        movq %rax, t2
        movq t2, %rax
        subq x, %rax
        movq %rax, %rax
\end{lstlisting}
  \hrule
  \caption{Example after Select Instructions, in the x86 assembly language
    augmented with variables.}
\label{fig:running-example-x86}
\end{figure}

@@

\noindent For the assignment statement we delegate to the
\lstinline{select-assign} function, which analyzes the expression
\lstinline{e} to determine which instruction to generate.
We describe \lstinline{select-assign} shortly.

@@

Next we look at some representative cases of
\lstinline{select-assign}. The first parameter is the right-hand side
of the assignment and the second parameter is the left-hand side, that
is, the destination. For subtraction, we use the \lstinline{rax}
register as scratch space. We first move argument \lstinline{a$₁$}
into \lstinline{rax}, then subtract \lstinline{a$₂$} from
\lstinline{rax}, and finally move the result into \lstinline{dest}.

@@

\subsubsection{Select Instructions is Correct}

@@

The correctness theorem delegates to the correctness lemma for
\lstinline{select-stmt}, which says that if statement \lstinline{c}
evaluates to \lstinline{v}, then the instructions generated for
\lstinline{c} put the same value \lstinline{v} into \lstinline{rax}.
The list of x86 instruction blocks is obtained by mapping
\lstinline{select-stmt} over the control-flow graph \lstinline{G}.

@@

\subsection{The Compiler is Correct}

@@

>>>>

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
%\documentclass[sigplan,screen,10pt]{acmart}
\documentclass[sigplan,screen]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{cc}
\setcctype{by}
\acmDOI{10.1145/3759427.3760372}
\acmYear{2025}
\copyrightyear{2025}
\acmISBN{979-8-4007-2150-2/25/10}
\acmConference[OLIVIERFEST '25]{Proceedings of the Workshop Dedicated to Olivier Danvy on the Occasion of His 64th Birthday}{October 12--18, 2025}{Singapore, Singapore}
\acmBooktitle{Proceedings of the Workshop Dedicated to Olivier Danvy on the Occasion of His 64th Birthday (OLIVIERFEST '25), October 12--18, 2025, Singapore, Singapore}
\received{2025-06-18}
\received[accepted]{2025-07-31}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Verified Nanopasses for Compiling Conditionals}
\titlenote{This work was partially funded by NSF award 1814460.}

@@

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011006.10011041</concept_id>
       <concept_desc>Software and its engineering~Compilers</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011074.10011099.10011692</concept_id>
       <concept_desc>Software and its engineering~Formal software verification</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10010124.10010131.10010134</concept_id>
       <concept_desc>Theory of computation~Operational semantics</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10010124.10010138.10010142</concept_id>
       <concept_desc>Theory of computation~Program verification</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Compilers}
\ccsdesc[500]{Software and its engineering~Formal software verification}
\ccsdesc[500]{Theory of computation~Operational semantics}
\ccsdesc[500]{Theory of computation~Program verification}

%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{short-cut boolean evaluation, nanopass, verified compilation, mechanized proof, Agda}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
%% \begin{teaserfigure}
%%   \includegraphics[width=\textwidth]{sampleteaser}
%%   \caption{Seattle Mariners at Spring Training, 2010.}
%%   \Description{Enjoying the baseball game from the third-base
%%   seats. Ichiro Suzuki preparing to bat.}
%%   \label{fig:teaser}
%% \end{teaserfigure}

@@

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to an algorithm of
\citet{Aho:1986qf} (presented in the form of an attribute grammar)
that translates Boolean expressions and conditionals into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarrassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
datatypes~\citep{PEYTONJONES19983}.

@@

Towards developing a course on verified compilation, I am interested
in proving correctness of the compiler presented in the
\emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. This textbook does not
include proofs of correctness for its compiler passes, though it is
organized in a way that might facilitate such proofs.
%
\emph{Essentials of Compilation} takes an \emph{incremental} approach
to teaching compilers, that is, it guides the students through the
creation of multiple compilers for a sequence of languages, each of
which adds a language feature to the previous language. For the
language with Boolean expressions and conditionals, the Explicate
Control pass of the compiler uses the one-function per context
approach of Danvy to translate into an x86-flavored register transfer
language.
%
In addition to being incremental, \emph{Essentials of Compilation}
employs the \emph{nanopass} approach, breaking up the compiler into
many small passes that are easier for students to
digest~\citep{Sarkar:2004fk}.
%
There are many other pedagogical treatments of
compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa}
but none of them take the incremental approach.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The Agda code is available at the following location.
\begin{center}
\url{https://doi.org/10.5281/zenodo.16890248}
\end{center}
The Agda code does not depend on special features of Agda and could
have been carried out in Rocq, Isabelle, or Lean.\footnote{The choice
of Agda for this paper is not an endorsement of Agda but is a
byproduct of the author's relative expertise in Agda at the time of
writing.}

@@

\subsection{The C-like Language}
\label{sec:c-lang}

@@

\subsection{X86 with Variables}

@@

\begin{figure}[tp]
  %\hrule
\begin{lstlisting}
    let x = read in
    let y = read in
    if (if x == 0
        then true
        else read == 1)
    then y - 1
    else (y - 2) - x
\end{lstlisting}
  %\hrule
\caption{An example source program.}
\label{fig:running-example-source}
\end{figure}

@@

\begin{figure}[tp]
  %\hrule
\begin{lstlisting}
    let x = read in
    let y = read in
    if (if x == 0
        then true
        else (let t1 = read in t1 == 1))
    then (y - 1)
    else (let t2 = y - 2 in t2 - x)
\end{lstlisting}
  %\hrule
\caption{\small Result of Remove Complex Operands on the example.}
\label{fig:running-example-monadic}
\end{figure}

@@

\paragraph{Remove Complex Operands Is Correct}

@@

\begin{figure}[tp]
  %\hrule
\begin{lstlisting}
    var x, y, t1, t2;
    x := read;
    y := read;
    if (if x == 0
        then true
        else (t1 := read; t1 == 1)
    then (y - 1)
    else (t2 := y - 2; t2 - x)
\end{lstlisting}
  %\hrule
  \caption{Result of Lift Locals on the example.}
  \label{fig:running-example-structured-imperative}
\end{figure}
The \lstinline{lift-locals} function delegates to the recursive
function \lstinline{lift-locals-mon}, where all the work is done.
That function returns (1) a number $n$ that counts the number of
\lstinline{let} bindings that were in the expression and (2) the
translation of the expression.

@@

\paragraph{Lift Locals Is Correct}

@@

\begin{figure}[tp]
  %\hrule
\begin{lstlisting}
    var x, y, t1, t2;
    block_0:
       x := read;
       y := read;
       if (x == 0) goto block_1; else goto block_3;
    block_1:
        goto block_2;
    block_2:
        return y - 1;
    block_3:
       t1 := read;
       if (t1 == 1) goto block_4; else goto block_5;
    block_4:
        goto block_2;
    block_5:
        goto block_6;
    block_6:
        t2 := y - 2;
        return t2 - x;
\end{lstlisting}
  %\hrule
  \caption{Example after Explicate Control, where control flow
    is made explicit with goto.}
\label{fig:running-example-goto}
\end{figure}

@@

For the finale of the Explicate Control pass, we discuss the case in
\lstinline{explicate-pred} for the conditional \lstinline{(If e$₁$ e$₂$ e$₃$)}.
The two branches \lstinline{e$₂$} and
\lstinline{e$₃$} are in predicate position, so we compile them with
recursive calls to \lstinline{explicate-pred} to obtain
\lstinline{c$₂$} and \lstinline{c$₃$}. However, they share the same
two continuations, \lstinline{thn} and \lstinline{els}, so we add both
of them to the CFG to avoid code duplication. We compile
\lstinline{e$₁$} with \lstinline{explicate-pred}, passing
\lstinline{c$₂$} and \lstinline{c$₃$} as the continuations.

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
explicate-pred (If e$₁$ e$₂$ e$₃$) thn els =
   add-node thn thenB
   $λ$ l$₁$ $→$ add-node els thenB
   $λ$ l$₂$ $→$ explicate-pred e$₂$ (Goto l$₁$) (Goto l$₂$) thenB
   $λ$ c$₂$ $→$ (explicate-pred e$₃$ (Goto l$₁$) (Goto l$₂$)) thenB
   $λ$ c$₃$ $→$ explicate-pred e$₁$ c$₂$ c$₃$
\end{lstlisting}

@@

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
explicate-pred-correct :
  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G G$′$}
  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
  $→$ ((s$′$,$ρ′$) , G $⊢ᶜ$ (if b then c$₂$ else c$₃$) $⇓$ v $⊣$ (s$″$,$ρ″$))
  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G $≡$ (c$₁$ , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
\end{lstlisting}

@@

\begin{figure}[tp]
  %\hrule
\begin{lstlisting}[mathescape=false,basicstyle=\ttfamily\footnotesize]
    var x, y, t1, t2
    block_0:
       callq read_int
       movq %rax, x
       callq read_int
       movq %rax, y
       cmpq x, $0
       je block_1
       jmp block_3
    block_1:
        jmp block_2
    block_2:
        movq y, %rax
        subq $1, %rax
        movq %rax, %rax
    block_3:
        callq read_int
        movq %rax, t1
        cmpq t1, $1
        je block_4
        jmp block_5
    block_4:
        jmp block_2
    block_5:
        jmp block_6
    block_6:
        movq y, %rax
        subq $2, %rax
        movq %rax, t2
        movq t2, %rax
        subq x, %rax
        movq %rax, %rax
\end{lstlisting}
  %\hrule
  \caption{Example after Select Instructions, in the x86 assembly language
    augmented with variables.}
\label{fig:running-example-x86}
\end{figure}

@@

\noindent For the assignment statement we delegate to
\lstinline{select-assign}, which analyzes the expression \lstinline{e}
to determine which instruction to generate.  We describe
\lstinline{select-assign} shortly.

@@

Next we look at representative cases of \lstinline{select-assign}. The
first parameter is the right-hand side of the assignment and the
second parameter is the left-hand side, that is, the destination. For
subtraction, we use the \lstinline{rax} register as scratch space. We
first move argument \lstinline{a$₁$} into \lstinline{rax}, then
subtract \lstinline{a$₂$} from \lstinline{rax}, and finally move the
result into \lstinline{dest}.

@@

\paragraph{Select Instructions Is Correct}

@@

The correctness theorem delegates to a lemma about the
\lstinline{select-stmt} function, which says that if statement
\lstinline{c} evaluates to \lstinline{v}, then the instructions
generated for \lstinline{c} put the same value \lstinline{v} into
\lstinline{rax}.  The list of x86 instruction blocks is obtained by
mapping \lstinline{select-stmt} over the control-flow graph
\lstinline{G}.

@@

\subsection{The Compiler Is Correct}

@@

\section{Data-Availability Statement}

The Agda files the define the compiler and proof of correctness are
available from \citet{Siek:2025aa}.

\balance
